Describing *Symbiodinium* metacommunities using high-throughput sequencing

### 2017-01-04 manuscript/manuscript.Rmd
update map figure

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and susceptibility to stress of their coral hosts. Therefore, describing the diversity and community structure of *Symbiodinium* assemblages in corals, as well as their metacommunities on the scale of whole reefs, has become a routine and essential goal in coral reef research. High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic vs. interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity. Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St. John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses variability that is more likely to be intragenomic, while maintaining variability that is more likely interspecific. While this approach cannot resolve all *Symbiodinium* species (e.g., when closely related species co-occur in a sample, or when ITS2 fails to differentiate them), it enables broad-scale metabarcoding of the evolutionarily complex, yet commonly-used ITS2 locus to generate taxa-abundance data that is suitable for ecological analysis. We utilized this approach to create a comprehensive metacommunity network of host-symbiont associations in St. John, which revealed *Symbiodinium* lineages that occupy 'dominant' vs. 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations, though no differences were observed between north and south shore environments. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

@@

**Be very explicit up front about what is meant by type vs species**

@@

  *Symbiodinium* are primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; Santos:2003kt; @Pochon:2012we]. Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on a combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes, of which the most commonly used is ITS2. High-throughput sequencing of ITS2 increasingly is being utilized to characterize these communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb]. However, such analyses create extremely large datasets consisting of millions of sequences, which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within this genus [@Finney:2010p8324; @Thornhill:2013dx]. For example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially other species [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (sensu XXX). In fact, concerted evolution may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and due to the high volume of existing sequence data for comparative analysis (e.g., GeoSymbio, etc.), and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms (XXX), it remains an essential and powerful marker for the *Symbiodinium* system. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.

  Here, we describe a *Symbiodinium* metacommunity with scleractinians and a *Millepora* hydrocoral in St. John, U.S. Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications by which such datasets may advance our understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' to extract ecologically relevant information from ITS2 metabarcoding data given its aforementioned limitations. Specifically, within-sample clustering addresses the fact that sequences within one species' genome may be only 97% similar [@Arif:2014gk], while the dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt]. Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates species diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity underestimates species diversity by collapsing species known to be different. True biological diversity in *Symbiodinium* may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts. These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to represent a single taxon rather than multiple, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa. Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using this within-sample clustering approach, we analyzed *Symbiodinium* communities in ten coral species across the north and south shores of St. John. We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) testing for differences in symbiosis flexibility (i.e., beta diversity) among coral species. The large quantity of data and the kinds of analytical approaches facilitated by ITS2 metabarcoding have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts. More generally, this work demonstrates how ecological knowledge can inform bioinformatic analyses given complex marker gene to species correspondence.

@@

#### Sample collection and processing

  A total of 87 tissue samples from scleractinian and milleporine corals were collected from seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and all colonies were selected haphazardly in upward facing locations exposed to ambient irradiance at the sampling depth. Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay. Samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB). Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.

#### Environmental data

@@

#### Bioinformatics

@@

#### Environmental and sequence data overview

@@

```{r Fig1, fig.height=3, fig.width=5, fig.cap="Figure 1. Sampling locations around St. John, U.S. Virgin Islands."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

@@

#pdf(file="figures/map.pdf", height=3, width=6)
layout(mat=matrix(c(1,1,1,2,1,1,1,3,1,1,1,4), nrow=3, byrow=T))
par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1)
plot(stjmap, lwd=1, col="gray40",
     xlim=c(-64.74, -64.68), ylim=c(18.303, 18.370))
#box()
text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.77, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.77, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black")
  text(lon, lat, name, pos=c(4,2,4,4,2,1,3))
})
points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
NS <- c(-64.711967, 18.368141)
SS <- c(-64.726840, 18.315465)
points(NS[1], NS[2], pch="*", cex=2, col="red")
points(SS[1], SS[2], pch="*", cex=2, col="red")
#dev.off()

@@

par(cex=0.7, mar=c(2,0,1,2), mgp=c(1.5,0.5,0), xpd=NA)
plot(NA, xlim=c(26,29), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="SST (°C)", xaxt="n")
axis(side=1, at=seq(26,29,1))
bxp(list(stats=matrix(as.numeric(sst[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
bxp(list(stats=matrix(as.numeric(sst[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)

@@

#par(new=T, mar=c(7,14,10,8), cex=0.7)
plot(NA, xlim=c(0.03,0.09), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="ChlA (mg m-3)", xaxt="n")
axis(side=1, at=seq(0.03,0.09,0.01))
bxp(list(stats=matrix(as.numeric(chla[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
bxp(list(stats=matrix(as.numeric(chla[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)

t(nsamp)[2,]
#dev.off()
```

@@

```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A.) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B.) 97% across-sample clustering and C.) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A.) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the closest BLAST hit to the representative sequence of the OTU; those with the same annotation but different fill colors represent unique sequence variants with the same closest BLAST hit."}
set.seed(42432)
par(mfrow=c(1,1))
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

@@

```{r Table3, cache=T} 
set.seed(43789) 
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 3. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")

@@

#### Network analyses of *Symbiodinium* meta-community
```{r networks,  cahce=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

@@

The observed structure of *Symbiodinium* communities found in a variety of host species was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data. Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset or independently within each sample affected the number and taxonomic assignment of OTUs. Clustering sequences from all samples together at 97% similarity reduced the total number of OTUs by 98% relative to no clustering (Table 1). Intermediate diversity was resolved by within-sample clustering, with 13% more OTUs than across-sample clustering. Comparison of the outcomes of these approaches (Fig. 2) demonstrates that across-sample clustering assigns the same dominant OTU to samples with dissimilar sequence assemblages. This results from the selection of an OTU's most abundant member sequence as its basis for taxonomic assignment: when member sequences occur across many samples (i.e., during across-sample clustering), the most abundant depends strongly on the number of samples of different hosts in the dataset. Consequently, closely-related sequences arising from different *Symbiodinium* types may be clustered into the same OTU which is then named according to the sequence variant that is most well-represented (i.e., most heavily sampled) in the dataset. Therefore, while clustering based on 97% similarity may successfully collapse intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when clustering is applied across a dataset potentially comprised of many closely related types, it may also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present.

  Meanwhile, clustering within samples assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the data set. The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa. Indeed, only within-sample clustering assigned different dominant clade C types to samples of *O. annularis* and *S. siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx]. Therefore, while multi-locus approaches may provide a superior description of *Symbiodinium* diversity, within-sample clustering of ITS2 metabarcoding data may be a good approximation of diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods exploit the numerically abundant sequence variants as taxonomically informative. Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa. Therefore, we recommend within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.

>>>>

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-susceptibility of their coral hosts. Therefore, describing the diversity and community structure of *Symbiodinium* assemblages in corals, as well as their metacommunities on the scale of whole reefs, has become a routine and essential goal in coral reef research. High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic versus interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity. Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St. John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses variability that is more likely to be intragenomic, while maintaining variability that is more likely interspecific. While this approach cannot resolve all *Symbiodinium* species (e.g., when closely related species co-occur in a sample, or when ITS2 fails to differentiate them), it enables broad-scale metabarcoding of the evolutionarily complex, yet commonly-used ITS2 locus to generate taxa-abundance data that is suitable for ecological analysis. We utilized this approach to create a comprehensive metacommunity network of host-symbiont associations in St. John, which revealed *Symbiodinium* lineages that occupy 'dominant' vs. 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations, though no differences were observed between north and south shore environments. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

@@

@@

  *Symbiodinium* are primarily described using genetic sequence data, including both chloroplast (cp23S and psbA^ncr^) and nuclear markers (18S, 28S, ITS1, and ITS2) [@LaJeunesse:2001p64; Santos:2003kt; @Pochon:2012we]. Together, these markers support nine major 'clades' within the genus *Symbiodinium* (clades A through I [@Pochon:2010p8784]), which are further subdivided into 'types' based on the marker used to identify them. *Symbiodinium* species are beginning to be described based on combinations of markers, including microsatellites to establish reproductive isolation (i.e., to satisfy the biological species concept) [@LaJeunesse:2014io]. However, until more species are described, ecological surveys of *Symbiodinium* diversity must rely on established marker genes, of which the most commonly used is ITS2. Consequently, high-throughput sequencing of ITS2 increasingly is being utilized to characterize these communities with unprecedented resolution [e.g., @Arif:2014gk; @Quigley:2014dv; @Thomas:2014da; @Edmunds:2014jc; @Cunning:2015hb; @Ziegler:2017cs]. However, such analyses create extremely large datasets consisting of millions of sequences, which places great importance on the ways in which ITS2 sequence diversity is analyzed and interpreted in relation to biological diversity.

  While ITS2 initially was investigated as a potential species-level marker in *Symbiodinium* [@LaJeunesse:2001p64], it is now understood that this marker is not sufficiently variable to distinguish all species within this genus [@Finney:2010p8324; @Thornhill:2013dx]. For example, the 'B1' ITS2 sequence variant is shared by *S. minutum*, *S. pseudominutum*, and *S. antillogorgium*, and potentially other species [@Parkinson:2015gt]. Moreover, the position of ITS2 within the tandemly repeating ribosomal DNA array results in the occurrence of multiple ITS2 sequence variants within a single genome [@Thornhill:2007p145] that evolve through concerted evolution (e.g. [@Dover:1986ju]). In fact, concerted evolution may mask species divergence by maintaining ancestral sequence variants as numerical dominants in multiple derived lineages [@Thornhill:2013dx]. Together, these features of ITS2 complicate the interpretation of intragenomic versus interspecific variation, and preclude its use as a true species-level marker. Nevertheless, numerically dominant intragenomic variants of ITS2 are still phylogenetically informative across the *Symbiodinium* genus [@LaJeunesse:2001p64], and resolve diversity at a functionally and ecologically important level. Moreover, due to the high volume of existing sequence data for comparative analysis (e.g., [@Franklin:2012jn; @Tonk:2013ey]), and the relative ease of amplifying and sequencing ITS2 using current high-throughput platforms, it remains an essential and powerful marker for the *Symbiodinium* system. Therefore, it is important to develop best practices in the bioinformatic analysis and interpretation of ITS2 metabarcoding surveys.

  Here, we describe a *Symbiodinium* metacommunity with scleractinians and a *Millepora* hydrocoral in St. John, U.S. Virgin Islands, with the objectives of: 1) developing an appropriate bioinformatic approach for *Symbiodinium* ITS2 metabarcoding surveys, and 2) exploring network analysis and statistical applications for such datasets that may advance our understanding of *Symbiodinium* metacommunity ecology. In addressing the first objective, we test the ability of 'within-sample clustering' to extract ecologically relevant information from ITS2 metabarcoding data given its aforementioned limitations. Specifically, within-sample clustering addresses the fact that sequences within one species' genome may be only 97% similar [@Arif:2014gk], while dominant intragenomic variants from different species may be >97% similar, or even identical [@Thornhill:2013dx; @Parkinson:2015gt]. Clearly, treating each ITS2 sequence as a unique *Symbiodinium* type overestimates species diversity due to the presence of intragenomic variants, while clustering sequences at 97% similarity may underestimate diversity by collapsing species known to be different. True biological diversity in *Symbiodinium* may be better approximated by a within-sample clustering approach that exploits key assumptions regarding the distribution of symbionts among samples, and sequences among symbionts. These assumptions include: 1) that most coral colonies are dominated by a single *Symbiodinium* type [@LaJeunesse:2011us; @Pettay:2011wt; @Baums:2014hj], and 2) that different numerically dominant ITS2 sequence variants diagnose different *Symbiodinium* types, even when they differ by only 1 nucleotide [@Sampayo:2009p3422]. These assumptions suggest that when closely related ITS2 sequences (i.e., that are >97% similar; @Arif:2014gk) occur within the same sample, they are more likely to represent a single taxon rather than multiple, but when they are numerically dominant in different samples, they are more likely to represent distinct *Symbiodinium* taxa. Accordingly, clustering sequences at 97% similarity within each sample independently may collapse variability that is more likely to be intragenomic (i.e., occurring within a sample), while maintaining variability that is more likely to be interspecific (i.e., occurring in different samples).

  Using this within-sample clustering approach, we analyzed *Symbiodinium* communities in ten coral species across the north and south shores of St. John. We explore ecological patterns that can be revealed by large-scale metabarcoding datasets, including 1) testing for whole-community differences associated with north and south shore locations, 2) analyzing coral-*Symbiodinium* metacommunity association networks, and 3) quantifying the variability in symbiont communities (i.e., beta diversity, or 'symbiosis flexibility') among individuals within a host species. The large quantity of data and the kinds of analytical approaches facilitated by ITS2 metabarcoding have the potential to revolutionize our understanding of *Symbiodinium* metacommunity ecology in reef corals and other taxa harboring similar symbionts. More generally, this work demonstrates how ecological knowledge can inform bioinformatic analyses given complex marker gene to species correspondence.

@@

#### Sample collection and environmental conditions

  A total of 87 tissue samples from scleractinian and milleporine corals were collected from seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and all colonies were selected haphazardly in upward facing locations exposed to ambient irradiance at the sampling depth. Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.

@@

#### ITS2 sequencing and bioinformatics

  Coral tissue samples were preserved in ~500 µL Guanidinium buffer (50% w/v guanidinium isothiocyanate; 50 mM Tris pH 7.6; 10 μM EDTA; 4.2% w/v sarkosyl; 2.1% v/v β-mercaptoethanol) and shipped to the Hawaii Institute of Marine Biology (HIMB). Genomic DNA was extracted from each coral tissue sample following a guanidinium-based extraction protocol [@Cunning:2015hb], and sent to Research and Testing Laboratory (Lubbock, TX) for sequencing of ITS2 amplicons ('itsD' and 'its2rev2' primers from @Stat:2009p7366) on the Illumina MiSeq platform with 2x300 paired-end read chemistry.

@@

#### Sample collection and environmental data

@@

```{r Fig1, fig.height=3, fig.width=6, fig.cap="Figure 1. Sampling locations around St. John, U.S. Virgin Islands."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

@@

pdf(file="figures/map.pdf", height=3, width=6)
layout(mat=matrix(c(1,1,6,2,1,1,5,3,1,1,5,4), nrow=3, byrow=T))

par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1, xpd=NA)
plot(stjmap, lwd=1, col="gray40",
     xlim=c(-64.75, -64.69), ylim=c(18.303, 18.370))
#box()
text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.77, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.77, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black")
  #text(lon, lat, name, pos=c(4,2,4,4,4,1,3))
})
#points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
#text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
NS <- c(-64.711967, 18.368141)
SS <- c(-64.726840, 18.315465)
points(NS[1], NS[2], pch="N", cex=0.9, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch="S", cex=0.9, bg="gray40", col="black", lwd=1)
points(NS[1], NS[2], pch=1, cex=2.2, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch=1, cex=2.2, bg="gray40", col="black", lwd=1)
#dev.off()

@@

par(cex=0.7, mar=c(2,0,1,2), mgp=c(1.5,0.5,0), xpd=NA)
plot(NA, xlim=c(26,29), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="SST (°C)", xaxt="n")
axis(side=1, at=seq(26,29,1))
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
bxp(list(stats=matrix(as.numeric(sst[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
bxp(list(stats=matrix(as.numeric(sst[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)

@@

#par(new=T, mar=c(7,14,10,8), cex=0.7)
plot(NA, xlim=c(0.03,0.09), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="ChlA (mg m-3)", xaxt="n")
axis(side=1, at=seq(0.03,0.09,0.01))
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
bxp(list(stats=matrix(as.numeric(chla[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
bxp(list(stats=matrix(as.numeric(chla[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)

plot(NA, xlim=c(0,1), ylim=c(0.5,2.5), xlab="", ylab="", bty="n", yaxt="n")
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
#text(par("usr")[1], c(1,2), c("O","O"), cex=1.8, pos=2, offset=0.7, lwd=0.1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)

#### histogram of samples
nsam <- with(samdat, aggregate(Species, by=list(Species=Species, Shore=Shore), FUN=length))
nsam <- dcast(nsam, Species~Shore)
rownames(nsam) <- nsam$Species

rownames(nsam) <- str_replace_all(rownames(nsam), 
                c("alcicornis"="M.alc.", "annularis"="O.ann.", "astreoides"="P.ast.","cavernosa"="M.cav.",
                  "cylindrus"="D.cyl.", "fragum"="F.fra.", "furcata"="P.fur.", "radians"="S.rad.",
                  "siderea"="S.sid.", "strigosa"="P.str."))
nsam <- nsam[,2:3]
par(mar=c(4,0,6,4))
barplot(t(as.matrix(nsam)), las=2)
legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1.5, legend=c("N","S"), bty="n", inset=c(0.1,-0.2), xpd=T, y.intersp=0.8, x.intersp=0.5)
dev.off()
```

@@

```{r Fig2, fig.width=6, fig.height=4, fig.align="center", fig.cap="Figure 2. Comparison of dominant OTUs assigned by 100%, 97% across-sample, and 97% within-sample clustering approaches. A.) The composition of unique sequence variants (i.e., 100% OTUs) in each sample, sorted by relative abundance, and filled by unique colors corresponding to unique sequence variants. The dominant OTUs within each sample assigned by B.) 97% across-sample clustering and C.) 97% within-sample clustering are shown by rectangles below each bar, with fill colors matching the unique sequence variants presented in (A.) to indicate the representative sequence of the assigned OTU. Colored rectangles that span multiple bars indicate that the corresponding samples were assigned the same dominant OTU. Rectangles are annotated with the taxonomic assignment of the OTU (see Methods)."}
set.seed(42432)
par(mfrow=c(1,1))
otubarplot2(samples=c("64","72","29","45","73","59","47","26","48","66","63"))
```

@@

```{r Table3, cache=T } 
set.seed(43789) 
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 3. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")

@@

#### Network analyses of *Symbiodinium* meta-community
```{r networks,  cache=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

@@

  The observed structure of *Symbiodinium* communities in a variety of host species was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data. Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset or independently within each sample affected the number and taxonomic assignment of OTUs. Clustering sequences from all samples together reduced the total number of OTUs by 98% relative to no clustering (Table 1), while within-sample clustering resolved intermediate diversity, with 13% more OTUs than across-sample clustering. Comparing taxonomic assignments of the OTUs generated by these approaches demonstrates that across-sample clustering in many cases assigns the same dominant OTU to samples with different dominant sequence variants (Fig. 2), while within-sample clustering assigns different OTUs that reflect the dominant variant within that particular sample. This occurs because an OTU's assigned identity is based on its most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it, with sequences from more heavily-sampled and/or deeply-sequenced hosts being better represented. The best-represented sequences then subsume closely-related but different sequences from other samples into OTUs that bear their names, despite the fact that these other samples may not contain the named sequences. For ITS2 OTUs, such merging across samples is undesirable since even single nucleotide differences in a sample's dominant ITS2 variant may indicate different *Symbiodinium* species. Therefore, even though clustering at 97% similarity successfully collapses intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related types, it is likely to also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present. Indeed, this limitation was also encountered by @Arif:2014gk when clustering closely-related clade C sequences across samples: despite being comprised mostly of C41 variants, ~3k sequences from *Acropora hemprichii* were subsumed into a single OTU that was named C1 after the dominant variant among the ~7k sequences from *Pocillopora verrucosa* with which they were clustered. These cases clearly demonstrate how clustering sequences across samples causes the identity of OTUs in any given sample to be driven largely by the other samples in the dataset.

  Meanwhile, clustering sequences independently within samples assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the dataset. The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa. Indeed, only the within-sample clustering approach assigned different dominant clade C OTUs to samples of *O. annularis* and *S. siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx]. Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically abundant sequence variants to assign taxonomy. Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa. Therefore, we recommend within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.
